---
title: "These are a few of my favorite things..."
description: |
  Some Functions and concepts I've learned so far
author:
  - name: Grace Lewin
    url: {}
date: 08-16-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

`summarize()` function:

The `summarize()` function creates a new data frame. This data frame will have rows for each combination of grouping variables. These grouping variables are specified by `group_by()`. The columns in this data frame are the summary statistics you have specified (ex. mean(), max(), min(), sd(), etc.)

Here is an example of using the `summarize()` function:

In this example I'm grouping by species, and island, and then finding the mean and standard deviation for flipper length (in mm) for those groups.

```{r}
mean_sd_penguins <- penguins %>% 
  group_by(species, island) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE))

mean_sd_penguins

```


